% ------------------------------------------------------------------
\newcommand{\thisunit}{MATH327 information}
\newcommand{\moddate}{Last modified 4 Feb.~2024}
\setcounter{section}{0}
\phantomsection
\addcontentsline{toc}{section}{Module information and logistics}
\section*{Module information and logistics}

\subsection*{Coordinator}
\begin{description}
  \setlength{\itemsep}{1pt}
  \setlength{\parskip}{0pt}
  \setlength{\parsep}{0pt}
  \item[\qquad] David Schaich
  \item[\qquad] Theoretical Physics Wing Room 123
  \item[\qquad] \href{mailto:david.schaich@liverpool.ac.uk}{david.schaich@liverpool.ac.uk}
  \item[\qquad] \href{https://calendly.com/daschaich/meet}{calendly.com/daschaich}
  \item[\qquad] \href{http://www.davidschaich.net}{www.davidschaich.net}
\end{description}
% ------------------------------------------------------------------



% ------------------------------------------------------------------
\subsection*{Overview}
Along with quantum mechanics and relativity, statistical physics is a central pillar of modern physics.
It uses stochastic (i.e., probabilistic) techniques to predict the large-scale behaviour that emerges from the microscopic dynamics of \emph{many} underlying objects --- for example, the $\sim$$10^{22}$ atoms in a cubic centimetre of water.
Applications of statistical physics range from nuclear physics and cosmology to climate science and biophysics, often with outstanding success.

The module outline on the previous page is organized around the concept of \textit{statistical ensembles} introduced in the early 1900s.
In essence, a statistical ensemble is a mathematical framework for concisely describing the properties of idealized physical systems subject to certain constraints.
After studying the probability foundations underlying these frameworks, we meet the \textit{micro-canonical ensemble} in unit 2 and the \textit{canonical ensemble} in unit 3.
The following two units 4--5 apply the canonical ensemble to investigate non-interacting (``ideal'') gases and thermodynamic cycles.
Unit 6 introduces our third and final statistical ensemble, the \textit{grand-canonical ensemble}, which units 7--8 apply to several types of non-interacting quantum gases.
Finally, in unit 9 we begin to explore the effects of interactions, which open up a much broader landscape of applications that we will survey for the remainder of the term.
% ------------------------------------------------------------------



% ------------------------------------------------------------------
\subsection*{Schedule}
Most weeks we will have the following schedule: \\[-30 pt]
\begin{itemize}
  \setlength{\itemsep}{1pt}
  \setlength{\parskip}{0pt}
  \setlength{\parsep}{0pt}
  \item Lecture at 9:00--10:00 on Tuesdays in Room 210
  \item Tutorial at 14:00–15:00 on Thursdays in Elec.~Eng.~Lecture Room E2 (202)
  \item Lecture at 11:00--13:00 on Fridays in Room 210
\end{itemize}
\ \\[-30 pt]
During weeks 2, 3 and 4 (starting 5, 12 and 19 February) we will instead have computer lab sessions to provide opportunities for you to work on the computer project discussed further below: \\[-30 pt]
\begin{itemize}
  \setlength{\itemsep}{1pt}
  \setlength{\parskip}{0pt}
  \setlength{\parsep}{0pt}
  \item Computer lab at 09:00--11:00 on Tuesdays in the Central Teaching Lab PC Centre [\href{https://www.liverpool.ac.uk/intranet/timetabling/online-room-catalogue/rooms/ctl-pctc-green.html}{Room 117, Green Zone}]
  \item Lecture at 11:00--13:00 on Fridays in Room 210
\end{itemize}

I will use Zoom to record lectures (and lecturey bits of tutorials and computer labs), which also provides a way for anyone off campus or out of town to connect remotely through \href{https://liverpool-ac-uk.zoom.us/j/91499188829?pwd=R1dPNjZQUDloS2pGVmxNdWFzenFLQT09}{this link} (meeting ID 914 9918 8829, passcode Math327!).
Zoom recordings will appear along with all other resources at \href{canvas.liverpool.ac.uk/courses/69036}{our Canvas site}, \\
\centerline{\href{https://canvas.liverpool.ac.uk/courses/69036}{canvas.liverpool.ac.uk/courses/69036}}

\textbf{Office hours} will take place after each scheduled activity.
Most weeks this will be at 10:00 on Tuesdays, 15:00 on Thursdays and 13:00 on Fridays.
During weeks 2, 3 and 4 this will instead be at 11:00 on Tuesdays and 13:00 on Fridays. 
In all cases office hours will be held in Room 123 of the Theoretical Physics Wing, and will also connect to the \href{https://liverpool-ac-uk.zoom.us/j/91499188829?pwd=R1dPNjZQUDloS2pGVmxNdWFzenFLQT09}{Zoom link} above.
If these times do not work with your schedule, you can also make an appointment through \href{https://calendly.com/daschaich/meet}{calendly.com/daschaich}, use the \href{https://canvas.liverpool.ac.uk/courses/69036/discussion_topics}{Canvas discussion board}, or reach me by email at \href{mailto:david.schaich@liverpool.ac.uk}{david.schaich@liverpool.ac.uk}.
I will aim to respond to emails and discussion board queries within 48 hours.
% ------------------------------------------------------------------



% ------------------------------------------------------------------
\subsection*{Assessment and academic integrity}
There will be three in-term assignments.
Each accounts for 15\% of the module mark, with the remaining 55\% coming from the final exam.
Although the deadlines listed below are not ideal, they have been centrally coordinated within the Department to minimize pile-up across different modules.
\begin{description}
  \item[15\%] A computer project due \textbf{Tuesday, 20 February}
  \item[30\%] Two equally weighted homework assignments, the first due \textbf{Tuesday, 5 March} and the second due \textbf{Friday, 26 April}
  \item[55\%] A two-hour in-person final examination to be centrally scheduled within the 20--31 May exam period
\end{description}

According to the University's \href{https://www.liverpool.ac.uk/media/livacuk/tqsd/code-of-practice-on-assessment/code_of_practice_on_assessment.pdf}{Code of Practice on Assessment} (COPA), late submissions completed within 120 hours after the submission deadline will have 5\% of the total marks deducted each 24-hour period after the deadline. % Section 6.2
Submissions more than five days late will be awarded zero marks, though I will still endeavour to provide feedback on them.
I will aim to return feedback and share model solutions within two weeks of the deadline.

For your other modules you already should have read and understood the \href{https://www.liverpool.ac.uk/media/livacuk/tqsd/code-of-practice-on-assessment/appendix_L_cop_assess.pdf}{Academic Integrity Policy} detailed in COPA Appendix L.
If you have any questions about what is or is not acceptable under this policy, please ask me or our Academic Integrity Officer Kamila Zychaluk.
In all cases, the spirit of the Academic Integrity Policy should take precedence over legalistic convolutions of the text.

In particular, I encourage you to discuss the in-term assignments with each other, since discussing and debating your work is a very effective way to learn.
Note that I say \textit{your work} --- your submissions for all assignments must be your own work representing your own understanding, and the examination must be done on your own.
It is unacceptable to copy solutions in part or in whole from other students (current or prior) or from other sources (commercial or otherwise).
Should you make use of resources beyond the module materials --- including generative AI tools such as ChatGPT --- these must be explicitly referenced in your submissions.
Clear and neat presentations of your workings and the logic behind them will contribute to your mark.
% ------------------------------------------------------------------



% ------------------------------------------------------------------
\subsection*{Main resources and materials}
The main materials we will use are the lecture notes you are currently reading.
As you read further, you will encounter gaps in the notes, which provide bite-sized exercises to help you check your understanding.
We will fill most gaps during lectures, but I encourage you to think about them for yourself.

The ten units into which the content is organized won't neatly match up with the twelve weeks of the term.
Some units will require more time than others.
Regular Canvas announcements will summarize what we cover each week.

We work in units where the Boltzmann constant $k = 1$, and logarithms have base $e$ unless otherwise specified (i.e., $\log x = \ln x$).
There is no need to memorize any equations.
Many equations are numbered so that they can be referenced later on, not necessarily because they are important.
Key results, definitions and concepts are highlighted by coloured boxes, and you should aim to be confident in your understanding of these.

These lecture notes were first written `live' during the 2021 and 2022 editions of this module.
While they are now much more stable, they continue to be improved, refined and sometimes corrected.
The ``Last modified'' date at the bottom of each page will flag any changes that occur during the term.
You can track the changes themselves through this version control repository: \\
\centerline{\href{https://github.com/daschaich/MATH327_2024}{github.com/daschaich/MATH327\_2024}}
This repository also provides tools for you to raise any issues you see.
The \href{https://software-carpentry.org}{Software Carpentry} project provides an introduction to \href{https://swcarpentry.github.io/git-novice/}{Version Control with Git}.

\subsubsection*{Expected background}
\textbf{No} prior exposure to quantum mechanics or computer programming is required --- all necessary information on these topics will be provided.
I do anticipate that you have previously seen the \href{https://en.wikipedia.org/wiki/Standard_deviation}{standard deviation}, the \href{https://en.wikipedia.org/wiki/Binomial_coefficient}{binomial coefficient}
\begin{equation*}
  \binom{N}{k} = \frac{N!}{k! \, (N - k)!} = \binom{N}{N - k}
\end{equation*}
that counts the number of possible ways to choose $k$ objects out of a set of $N \geq k$ total objects, and \href{https://en.wikipedia.org/wiki/Gaussian_integral}{gaussian integrals},
\begin{align*} % TODO: a > 0...
  \int_{-\infty}^{\infty} e^{-a (x + b)^2} \; dx & = \sqrt{\frac{\pi}{a}} &
  a & > 0.
\end{align*}

\subsubsection*{Programming}
You are welcome to complete the computer project using the programming language of your choice.
I recommend \href{https://www.python.org}{Python}, which is free, user-friendly, and very widely used around the world.
During the first two weeks of the term we will review \href{https://tinyurl.com/math327demo}{this demo} that explains all the Python programming tools you'll need.
Python may already be available on devices you have access to --- either a personal computer or computers in our library.
If it isn't, you can also run Python code on many Web sites, of which I have had the best experiences with Google's \href{https://colab.research.google.com}{Colaboratory} and \href{https://cocalc.com}{CoCalc}. % replit.com got very slow in 2022; onlinegdb.com doesn't seem to handle figure display or saving; mybinder.org requires git repo; trinket.io sometimes struggles with longer calculations or plotting histograms
You may need to create a free account, and you should make sure to save a local copy to reduce the risk of losing your work.
Alternative languages could include \href{https://en.wikipedia.org/wiki/C_(programming_language)}{C}, \href{https://fortran-lang.org}{Fortran}, \href{https://www.r-project.org}{R}, or even \href{https://matlab.mathworks.com}{MATLAB} (through the University's site license).
Maple may struggle to handle parts of the project.
% ------------------------------------------------------------------



% ------------------------------------------------------------------
\subsection*{How to get the most out of this module}
At this point in your studies, this advice should be familiar, but it's worth repeating.

Come to class --- ideally in person, if necessary by Zoom.
This will ensure regular contact with the material, and help you check that you understand it.
If the module is moving slower than you'd prefer, coming to class will give you opportunities to ask about more interesting extensions, applications or complications.

Before class, take a quick look at the upcoming pages in the lecture notes, and think about how any gaps could be filled.
Look for the big ideas rather than digging in to every detail, and see if you have any questions (or objections) to raise in class.
After class, take a closer look at the details, and make sure the gaps have been filled to your satisfaction.
Even though the lecture notes reflect my plans for the module, they may not exactly match what happens in class, especially when questions arise.
We may gloss over some topics that seem to be sufficiently well explained in the notes, and we may delve deeper into other topics that seem to merit further consideration.

Work on the homework problems, computer project and tutorial activities.
The best way to learn mathematics is by doing mathematics, and these assignments and activities are designed to make you think and help develop your mathematical muscles.
In particular, the homework problems will be harder than exam questions, since you'll have much more time to work on them --- so make sure you start thinking about them well in advance of the deadline.
Afterwards, review the model solutions and feedback, to make sure any confusing points are resolved.

Ask questions.
Ask questions you think you're supposed to know the answer to.
Ask questions you think everyone else knows the answer to.
(They don't.)
Ask questions about the big ideas, the specific details, and the connections between them.
The opportunity to ask questions is the main benefit of taking a module.
You can ask me; you can ask your classmates; you can ask the additional resources below.
% If you ask me about assignments, I will avoid doing the work for you; instead I will have you explain to me what you have tried so far, and will ask leading questions to suggest where I see problems or potential next steps.
% ------------------------------------------------------------------


% ------------------------------------------------------------------
\subsection*{Additional resources}
The lists below provide some optional additional resources that may be helpful.
You can use the module \href{https://canvas.liverpool.ac.uk/courses/69036/external_tools/102}{Reading List} on Canvas to reach our library's records for the books.

\noindent\textbf{Resources at roughly the level of this module:} \\[-24 pt]
\begin{enumerate}
  \item David Tong, \href{https://www.damtp.cam.ac.uk/user/tong/statphys.html}{\textit{Lectures on Statistical Physics}} (2012), \\ www.damtp.cam.ac.uk/user/tong/statphys.html
  \item MIT OpenCourseWare for undergraduate \href{https://ocw.mit.edu/courses/8-044-statistical-physics-i-spring-2013/}{Statistical Physics I} (2013) and \href{https://ocw.mit.edu/courses/8-08-statistical-physics-ii-spring-2005/}{Statistical Physics II} (2005), \\ ocw.mit.edu/courses/8-044-statistical-physics-i-spring-2013/ \\ ocw.mit.edu/courses/8-08-statistical-physics-ii-spring-2005/
  \item Daniel V.~Schroeder, \textit{An Introduction to Thermal Physics} (2021)
  \item C.~Kittel and H.~Kroemer, \textit{Thermal Physics} (1980)
  \item F.~Reif, \textit{Fundamentals of Statistical and Thermal Physics} (1965)
\end{enumerate}

\noindent\textbf{More advanced and more specialized resources}, which may be useful to consult concerning specific questions or topics: \\[-24 pt]
\begin{enumerate}
  \setcounter{enumi}{5}
  \item MIT OpenCourseWare for postgraduate \href{https://ocw.mit.edu/courses/8-333-statistical-mechanics-i-statistical-mechanics-of-particles-fall-2013/}{Statistical Mechanics I} (2013) and \href{https://ocw.mit.edu/courses/8-334-statistical-mechanics-ii-statistical-physics-of-fields-spring-2014/}{Statistical Mechanics II} (2014), \\ ocw.mit.edu/courses/8-333-statistical-mechanics-i-statistical-mechanics-of- \\ particles-fall-2013/ \\ ocw.mit.edu/courses/8-334-statistical-mechanics-ii-statistical-physics-of- \\ fields-spring-2014/
  \item R.~K.~Pathria, \textit{Statistical Mechanics} (1996)
  \item Sidney Redner, \textit{A Guide to First-Passage Processes} (2007)
  \item Pavel L.~Krapivsky, Sidney Redner and Eli Ben-Naim, \textit{A Kinetic View of Statistical Physics} (2010)
  \item Kerson Huang, \textit{Statistical Mechanics} (1987)
  \item Andreas Wipf, \textit{Statistical Approach to Quantum Field Theory} (2021)
  \item Weinan E, Tiejun Li and Eric Vanden-Eijnden, \textit{Applied Stochastic Analysis} (2019)
  \item Michael Plischke \& Birger Bergersen, \textit{Equilibrium Statistical Physics} (2005)
  \item Sacha Friedli and Yvan Velenik, \textit{Statistical Mechanics of Lattice Systems} (2018)
  \item L.~D.~Landau and E.~M.~Lifshitz, \textit{Statistical Physics, Part 1} (1980)
\end{enumerate}

\noindent\textbf{A general book about learning}, emphasizing (among other things) the value of retrieval practice compared to re-reading lecture notes or re-watching videos: \\[-24 pt]
\begin{enumerate}
  \setcounter{enumi}{15}
  \item Peter C.~Brown, Henry L.~Roediger III and Mark A.~McDaniel, \textit{Make it Stick: The Science of Successful Learning} (2014) \\
        A \href{https://www.youtube.com/watch?v=MfylloWuuZU}{short summary video} is also available
\end{enumerate}

\noindent\textbf{Programming resources:} \\[-24 pt]
\begin{enumerate}
  \setcounter{enumi}{16}
  \item MATH327 \href{https://tinyurl.com/math327demo}{Python programming demo} (2024)
  \item \href{https://wiki.python.org/moin/BeginnersGuide}{Beginner's Guide to Python} (2023)
  \item \href{https://software-carpentry.org}{Software Carpentry} tutorials: \\
        \textcolor{white}{hack} \href{https://swcarpentry.github.io/git-novice/}{Version Control with Git} (2023) \\
        \textcolor{white}{hack} \href{https://swcarpentry.github.io/python-novice-inflammation/}{Programming with Python} (2024) \\
        \textcolor{white}{hack} \href{https://swcarpentry.github.io/python-novice-gapminder/}{Plotting and Programming in Python} (2023)
  \item Stormy Attaway, \textit{MATLAB: A Practical Introduction to Programming and Problem Solving} (2013)
  \item B.~Barnes and G.~R.~Fulford, \textit{Mathematical Modelling with Case Studies: Using Maple and MATLAB} (2014)
\end{enumerate}

In addition, there is a vast constellation of other online resources such as \href{https://physics.stackexchange.com/questions/tagged/statistical-mechanics}{StackExchange} and \href{https://en.wikipedia.org/wiki/Statistical_physics}{Wikipedia}.
These can be great places to \emph{start} learning about a topic, but are often terrible places to \emph{stop}.
% ------------------------------------------------------------------
